A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-19 podc_paper/algorithm.tex
working on algorithm description

We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each node of the tree stores an array of \emph{blocks} and each \emph{block} represents a 
sequence of enqueues and a sequence of dequeues.
To perform an operation on the queue, a process $P$ creates a block containing that single 
operation and appends it to the array in $P$'s leaf.
Then, $P$ attempts to propagate the operation along the path from that leaf to the root of the tree.
At each node $v$ along this path, $P$ combines information from the blocks in $v$'s children into a single
block and attempts to append this block to $v$'s array using a CAS instruction.
This cooperative approach is sufficient to ensure that if $P$ fails to CAS a new block into $v$'s array twice,
then its operation has been propagated to $v$ by some other process.
We define a total order on all operations that have been propagated to the root, which is used to
linearize the operations.
In this section, we assume for simplicity that each node has an infinite blocks array.
We describe in Section \ref{??} how to replace the infinite array by a representation that uses bounded space.

If $P$'s operation is an enqueue, it can terminate when the operation has been propagated to the root.
If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
(since the operation may have been propagated to the root by some other process).
Then, $P$ determines whether the queue is empty when its dequeue is linearized. If so, it returns \nil.
If not, $P$ computes the rank $r$ of its dequeue in the linearization ordering
among all dequeues that do not return \nil,
finds the $r$th enqueue in the linearization and returns that enqueue's value.

The primary challenge is thus figuring out what information to store in each block so that 
the following tasks can be done efficiently (in a polylogarithmic number of steps).
\begin{itemize}
\item
Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children.
\item
Given an operation in a leaf that has been propagated to the root, find that operation's location in the root's blocks array.
\item
Given the location of a dequeue operation in the root's blocks array, find the enqueue whose value it should return (or determine that the queue is empty when the dequeue is linearized).
\end{itemize}

To propagate operations to a node $n$ in the tree, a process observes
the operations in both of $n$'s children that are not already in $n$,
merges them to create an ordering and then tries to append the
ordering to the sequence stored in $n$. We call this procedure
\tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on
$n$ with a successful append helps to propagate their operations up to
$n$. We shall prove that if a process invokes \tt{Refresh} on the node
$n$ two times and fails to append the new operations to $n$ both
times, the operations that were in $n$'s children before the first
\tt{Refresh} are guaranteed to be in $n$ after the second failed
\tt{Refresh}. 
We sketch the argument here.

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{After the \nf{Refresh}.}
\end{subfigure}
\hspace{6em}
\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before
  and after a \nf{$n$.Refresh} with a successful append. Operations
  propagating from the left child are labelled with $l$ and from the
  right child with $r$.} 
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
  \caption[Two consecutive failed \nf{Refresh}es by a
    process.]{\label{fig::simpleDoubleRefresh}Time relations between
    the concurrent successful \nf{Refresh}es and the two consecutive
    failed \nf{Refresh}es.} 
\end{figure}

We use \nf{CAS} (Compare\&Swap) instructions to implement the
\nf{Refresh}'s attempt to append  described in the previous
paragraph. 
The second failed \nf{Refresh} of $P$ is assuredly concurrent with a
successful \texttt{Refresh} that has read its information after the
invocation of the first failed \nf{Refresh} (see Figure
\ref{fig::simpleDoubleRefresh}). This is because some process $L$ does
a successful append during $P$'s first failed attempt, and some
process $K$ performs a \nf{Refresh} that reads its information after
$L$'s append and then performs a successful append during $P$'s second
failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the
new operations that were in $n$'s children before $P$'s first failed
\nf{Refresh}, in case they were not already appended. After a process
appends its operation into its leaf it can call \nf{Refresh} on the
path up to the root at most two times on each node. So, with $O(\log
p)$ \nf{CAS}es an operation can ensure it appears in the
linearization. This cooperative solution allows us to overcome the CAS
Retry Problem.

It is not efficient to explicitly store the sequence of operations in
each node because each operation would have to be copied all the way
up to the root; doing this would not be possible in poly-logarithmic
time. 
Instead we use an implicit representation of the operations propagated
together. Furthermore, we do not need to maintain an ordering on
operations propagated together in a node until they have reached the
root. It is sufficient to only keep track of sets of operations
propagated together in each \nf{Refresh} and then define the
linearization ordering only in the root (see Figure
\ref{fig::set}). Achieving a constant-sized implicit representation of
operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions
on fixed-size objects in each \tt{Refresh}. To do that, we introduce
\it{block}s. A block stores information about the operations
propagated by a \nf{Refresh}. It contains the number of operations
from the left and the right child propagated to the node by the
\texttt{Refresh} procedure. See Figure~\ref{fig:block} for an
example. A node stores an array of blocks of operations propagated up
to it. A propagate step  aggregates the new blocks in the children
into a new block and stores the new block in the parent. We call the
aggregated blocks \it{subblocks} of the new block and the new block
the \it{superblock} of them.  In each \nf{Refresh} there is at most
one operation from each process trying to be propagated, because one
process cannot invoke two operations concurrently. Thus, there are at
most $p$ operations in a block. Furthermore, since the operations in a
\texttt{Refresh} step are concurrent we can linearize them  among
themselves in any order we wish, because if two operations are read in
one successful \tt{Refresh} step in a node they are going to be
propagated up to the root together. Our choice is to put the
operations propagated from the left child before the operations
propagated from the right child. In this way, if we know the number of
operations from the left child and the number of operations from the
right child in a block, we have a complete ordering of the
operations.

@@

So far, we have a shared tree that processes use to agree on the
implicit ordering stored in its root. With this agreement on the
linearization ordering, we can design a universal construction; for a
given object, we can perform an operation $op$ by applying all the
operations up until $op$ in the root on a local copy of the object and
then returning the response for $op$. However, this approach is not
enough for an efficient queue. 
We show that we can build an efficient queue if we can compute two
things about the ordering in the root: (1) the $i$th propagated
operation and (2) the rank of a propagated operation in the
linearization. We explain how to implement (1) and (2) in
poly-logarithmic steps.

After propagating an operation \textit{op} to the root, processes can
find out information about the linearization ordering using (1) and
(2).  
To get the $i$th operation in the root, we find the block $B$
containing the $i$th operation in the root, and then recursively find
the subblock of $B$ in the descendent of the root that contains that
$i$th operation. When we reach a block in a leaf, the operation is
explicitly stored there. To make this search faster, instead of
iterating over all blocks in the node, we store the prefix sum of the
number of elements in the blocks sequence to permit a binary search
for the required block. We also store pointers to determine the range
of subblocks of a block to make the binary search faster. In each
block, we store the prefix sum of operations from the left child and
from the right child. Moreover, for each block, we store two
attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of
the last left and right subblock (see Figure \ref{fig::pointer}). We
know a block size is at most $p$, so binary search takes at most
\textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and
\nf{end\sub{right}} indices of a block and its previous block reduce
the search range size to \textsc{O}$(p)$.

>>>>

\subsection{Overview of Implementation}
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
In this section, we assume for simplicity that each node has an infinite blocks array.
Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.

To perform an operation on the queue, a process $P$ creates a block containing that single 
operation and appends it to the blocks array in $P$'s leaf.
Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
We shall define a total order on all operations that have been propagated to the root, which is used to
linearize the operations.

To propagate an operation to a node $v$ in the tree, $P$ first observes
the blocks in both of $v$'s children that are not already in $v$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $v$'s blocks array using a CAS instruction.
Following \cite{}, we call this a 3-step sequence a
\op{refresh} on $v$. %(see Figure \ref{fig::propagstep}).
A successful \op{refresh} by $P$ may propagate pending operations by several other processes to $v$.
This cooperative approach is sufficient to ensure that if $P$ fails to CAS a 
new block into $v$'s array twice,
then its operation has been propagated to $v$ by some other process, so $P$ can continue 
onwards towards the root.

Now suppose $P$'s operation has been propagated all the way to the root.
If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
(since the operation may have been propagated to the root by some other process).
Then, $P$ determines whether the queue is empty when its dequeue is linearized. If so, it returns \nil.
If not, $P$ computes the rank $r$ of its dequeue in the linearization ordering
among all dequeues that do not return \nil,
finds the $r$th enqueue in the linearization and returns that enqueue's value.

The primary challenge is thus figuring out what information to store in each block so that 
the following tasks can be done efficiently (in a polylogarithmic number of steps).
\begin{enumerate}[label={(T\arabic*)}]
\item
\label{construct}
Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children, as required for a \op{refresh}.
\item
\label{findinroot}
Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's blocks array.
\item
\label{findrank}
Given a dequeue's position in the root, determine that the queue is empty when it occurs
or determine the rank $r$ of the enqueue whose value it should return.
\item
\label{findenqueue}
Find the $r$th enqueue in the linearization ordering.
\end{enumerate}
Since these tasks require knowing the linearization ordering, we next describe that ordering.

\subsection{Linearization Ordering}

An operation can terminate only after a block containing it has been appended to the root's blocks array.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
Thus, our linearization orders operations according to the block they belong to in the root's blocks array.
Operations that appear in the same block are necessarily concurrent, so we can choose how to order them.

Each block in a leaf represents a single operation.
Each block $B$ in an internal node $v$ results from merging
several consecutive blocks from each of $v$'s children.
The blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
or a subblock of a direct subblock of $B$.
A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.

The operations propagated by a \op{refresh} are all pending when the \op{refresh} occurs,
so there is at most one operation per process.
Hence, a block represents at most $p$ operations in total.  
Moreover, we never append empty blocks, so 
each block represents at least one operation and it follows that a block can have at most $p$ direct subblocks.

As mentioned above, we are free to order operations within a block however we like.
For convenience, we order the enqueues and dequeues separately, and put the 
operations that were propagated from the left child before the operations from the right child.
More formally, we inductively define an sequence $E(B)$ of the enqueues represented by a block $B$.
If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is either the single
enqueue represented by the block (if $B$ stores an enqueue), or the empty sequence (if $B$ stores a dequeue).
If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s enqueue sequence is the concatenation $E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)$.
The dequeue sequence $D(B)$ of a block $B$ is defined symmetrically.

Finally, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
linearization ordering is 
$E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k)$.

\here{Need some figures illustrating tournament tree, ordering of operations, representation as blocks.  Might be able to modify Fig \ref{fig::set} and \ref{fig:block}.}

@@

\subsection{Representing Blocks}

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block (task \ref{construct}), which would be much too slow.
Instead, the block stores an implicit representation of the sequences.

After propagating an operation \textit{op} to the root, processes can
find out information about the linearization ordering using \ref{findenqueue} and
\ref{findinroot}.  
To get the $i$th operation in the root, we find the block $B$
containing the $i$th operation in the root, and then recursively find
the subblock of $B$ in the descendent of the root that contains that
$i$th operation. When we reach a block in a leaf, the operation is
explicitly stored there. To make this search faster, instead of
iterating over all blocks in the node, we store the prefix sum of the
number of elements in the blocks sequence to permit a binary search
for the required block. We also store pointers to determine the range
of subblocks of a block to make the binary search faster. In each
block, we store the prefix sum of operations from the left child and
from the right child. Moreover, for each block, we store two
attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of
the last left and right subblock (see Figure \ref{fig::pointer}). We
know a block size is at most $p$, so binary search takes at most
\textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and
\nf{end\sub{right}} indices of a block and its previous block reduce
the search range size to \textsc{O}$(p)$.
